Inferring disease architecture and predictive ability\\with LDpred2-auto

### 2022-09-08 paper/paper-infer.tex
finished draft

Most traits and diseases are heritable, if not all. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Knowing how much a phenotype is heritable and how much it is polygenic can teach a lot about it.
Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,zeng2021widespread,johnson2021estimation}.

@@

Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
We call this new preferred set of 1,444,196 SNPs HapMap3+.

@@

\begin{figure}[!h]
	\centerline{\includegraphics[width=0.6\textwidth]{overview}}
	\caption{[TODO: ADD AN OVERVIEW FIGURE OF WHAT LDPRED2 CAN PROVIDE?] \label{fig:overview}}
\end{figure}

@@

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.

@@

Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
However, we note that we are able to achieve an $r^2$ of 0.67 [TODO: UPDATE NUMBER] when using the penalized regression implementation of \cite{prive2019efficient} while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.

@@

Nevertheless, LDpred2-auto users can now get much from running a single method only. The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions. 
As future research directions, we are interested in using it for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, and multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}.

>>>>

Most traits and diseases are heritable, if not all. 
What differs is the proportion of phenotypic variance that can be attributable to genetics, i.e.\ the heritability of these phenotypes.
Some phenotypes, such as schizophrenia or height, are highly heritable \cite[]{sullivan2003schizophrenia,yang2010common}. 
These two phenotypes are also highly polygenic, i.e.\ mutations from many genetic variants influences them \cite[]{oconnor2019extreme}.
Knowing how much a phenotype is heritable and how much it is polygenic can teach a lot about it.
Therefore, many methods have been developed to estimate the (SNP) heritability $h^2$ and polygenicity $p$.
These include GCTA ($h^2$), BOLT-REML ($h^2$ and $p$), LD Score regression ($h^2$), HESS (local $h^2$), LDAK-SumHer ($h^2$), S-LD4M ($p$), GRM-MAF-LD ($\alpha$ that can inform about negative selection), SuSiE (per-variant $p$), SBayesS ($h^2$, $p$, and a third parameter $S$, similar to $\alpha$), and BEAVR (local $p$) \cite[]{yang2011gcta,loh2015contrasting,bulik2015ld,shi2016contrasting,speed2019sumher,oconnor2019extreme,schoech2019quantification,wang2020simple,zeng2021widespread,johnson2021estimation}.

@@

Here we extend LDpred2-auto to make it a highly reliable method for estimating (local) $h^2$, (per-variant) $p$, and $\alpha$.
So, on top of providing competitive PGS, LDpred2-auto can now provide all these estimates of genetic architecture. 
Moreover, we show how it can now also reliably estimate the predictive ability $r^2$ of PGS it derives, allowing for directly assessing the usefulness of the derived PGS, without needing an independent test set.
An overview of what LDpred2-auto can provide is presented in Figure \ref{fig:overview}.
Finally, we extend the set of HapMap3 variants recommended to use with LDpred2, which enables us to capture around 12\% more SNP heritability and achieve around 6\% more predictive performance $r^2$.
We call this new preferred set of 1,444,196 SNPs HapMap3+.

@@

\begin{figure}[!htb]
	\centerline{\includegraphics[width=0.8\textwidth]{overview}}
	\caption{Overview of what LDpred2-auto can provide. For individual CIs of polygenic scores, please refer to the work of \cite{ding2022large}. \label{fig:overview}}
\end{figure}

@@

Across 169 UKBB phenotypes with more than 25 chains kept, we compute the median heritability per block, and compare it to the number of variants in these blocks; the median heritability explained by a block of variants is largely proportional to the number of variants in this block (Figure \ref{fig:median_local_h2}).
The outlier block explaining a much larger heritability contains the HLA region.
Across the same phenotypes, we then compute per-variant median probabilities of being causal, and report them in a Manhattan-like plot in Figure \ref{median_postp}.
Some variants in multiple small regions across the genome have a larger probability of being causal across many phenotypes; these are mapped to genes that are known to be associated with many different traits (up to more than 300) in the GWAS Catalog \cite[]{buniello2019nhgri}.
To verify that this is not driven by population structure, we compute pcadapt chi-square statistics that quantify whether a variant is associated with population structure \cite[]{prive2020performing}; the log-statistics have a negative correlation of -5.5\% with the probabilities of being causal.
To verify that this does not correspond to regions of low LD, we compute LD scores; the probabilities above have a correlation of 11.6\% with the LD scores and of -6.8\% with the log of LD scores.

@@

Here we have also extended the set of HapMap3 variants recommended to use with LDpred2, making it 40\% larger to offer a better coverage of the genome.
This enables us to capture more of the heritability of phenotypes and therefore reduce the missing heritability (i.e.\ the difference between the family-based heritability and the SNP-based heritability).
Using this new set also improves predictive performance here, particularly for lipoprotein(a) concentration with an $r^2$ of 0.516 instead of 0.344.
However, we note that we are able to achieve an $r^2$ of 0.677 [0.671, 0.682] when using the penalized regression implementation of \cite{prive2019efficient} while restricting to all variants within a 1Mb window of the \textit{LPA} gene.
It means that this extended SNP set is still not tagging all variants perfectly, and that it might be preferable to use a more targeted set of variants for phenotypes for which most of the heritability is contained in a single region of the genome.

@@

Nevertheless, LDpred2-auto users can now get much from running a single method only. The reliable estimates provided by LDpred2-auto are very encouraging to further extend LDpred2-auto in multiple directions. 
As future research directions, we are interested in using it for GWAS summary statistics imputation \cite[]{rueger2018evaluation,julienne2019raiss}, for genetic correlation estimation \cite[]{bulik2015atlas,shi2017local,speed2019sumher,frei2019bivariate,werme2022integrated}, multi-ancestry prediction and inference \cite[]{brown2016transethnic,shi2020localizing,ruan2022improving,lu2022multi}, as well as extending it to learn from functional annotations \cite[]{zhang2021improved,marquez2021incorporating}.
